---
title: "Numerical Data"
description: "Descriptive stats & visualizations for quantitative variables (mean/median, spread, hist/box/scatter)"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    theme: cosmo
    smooth-scroll: true
editor: source
knitr:
  opts_chunk: 
    echo: true
    eval: true
    message: true
    error: true
    warnings: false
    comment: "#>" 
execute:
  cache: true
---

# Setup

```{r}
#| label: setup
#| include: false
# Load packages used in this handout.
# If a package is missing on your machine, uncomment the install.packages lines
# and run them once. (Your instructor may have these preinstalled.)

# install.packages(c("tidyverse", "DT"))
# install.packages("csucistats")
# install.packages("patchwork")   # for multi-plot examples (optional)

library(tidyverse)
library(csucistats)
theme_set(theme_bw())
```

# Learning goals

- Identify **numerical (quantitative)** variables and read their distributions.
- Compute core descriptive statistics: **min, Q1, median (Q2), mean, Q3, max, IQR, variance, sd**.
- Make and interpret **histograms**, **box plots**, **dot plots**, and **scatter plots**.
- Recognize **outliers** and what they may indicate.
- Use copy‑paste **templates** and replace placeholders with your data/columns.

::: callout-tip
Use the **Copy** button on each code chunk. Most topics have a **template** first and a **worked example** using the *Mr. Trash Wheel* dataset.
:::

# Google Colab Setup

Copy the following code and put it in a code cell in Google Colab. Only do this if you are using a completely new notebook.

```{r}
#| eval: false
# This code will load the R packages we will use
install.packages(c("csucistats"),
                 repos = c("https://inqs909.r-universe.dev", "https://cloud.r-project.org"))
library(csucistats)
library(tidyverse)

# Uncomment and run for categorical plots
# csucistats::install_plots()
# library(ggtricks)
# library(ggmosaic)
# library(waffle)

# Uncomment and run for themes
# csucistats::install_themes()
# library(ThemePark)
# library(ggthemes)
```


## Data for this handout

We will use the **Mr. Trash Wheel** dataset from TidyTuesday.

```{r}
#| label: load-data
#| code-fold: true
#| message: false
trashwheel <- read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-03-05/trashwheel.csv")
```

```{r}
#| echo: false
DT::datatable(dplyr::slice_sample(trashwheel, n = 10), options = list(dom = "p", pageLength = 5))
```

## Using the templates: what to change

Use this legend whenever you see a **Template** code block.

- **`DATA`** → replace with your data frame/tibble name (e.g., `trashwheel`).

- **`VAR`** → replace with the **single numerical variable** you want (e.g., `PlasticBottles`).

  - In `ggplot(DATA, aes(x = VAR))`, write `ggplot(trashwheel, aes(x = PlasticBottles))`.

  - In functions that take a vector (e.g., `mean(DATA$VAR)`), write `mean(trashwheel$PlasticBottles)`.

- **`VAR1`** and **`VAR2`** → replace with the **x** and **y** variables for scatter plots (e.g., `PlasticBottles`, `PlasticBags`).

- **`bins = X`**, **`binwidth = X`** → choose a sensible number/width for your data scale.


### Quick replace checklist
1. Swap **`DATA`** for your data frame (usually `trashwheel`).

2. Swap **`VAR`** for your numerical column (e.g., `PlasticBottles`).

3. For scatter plots, set **`VAR1`** and **`VAR2`**.

4. Adjust **`bins`** or **`binwidth`** for histograms/dot plots to show the distribution clearly.


# Summary statistics

## What is numerical data?

Numerical (quantitative) variables record numbers for which arithmetic makes sense (e.g., items collected, weights, costs).

**Example (first few values):**
```{r}
head(trashwheel$PlasticBottles)
```

## Central tendency

**Central tendency** summarizes a distribution with a representative value (mean or median).

- **Median (Q2)** is the 50th percentile: half the data lie below it.

- **Mean** is the arithmetic average: sensitive to outliers/skew.


## Variation (spread)

**Variation** describes how far data tend to fall from the center.

- **Range** = max − min

- **IQR** = Q3 − Q1 (middle 50%)

- **Variance/SD** measure average squared/root‑mean‑squared distance from the mean.


## Summary Statistics

**Template:**

```r
num_stats(DATA$VAR)           # five-number summary + mean
```

**Example:**

```{r}
num_stats(trashwheel$PlasticBottles)
```

**Alternative:**

```r
summary(DATA$VAR)           # five-number summary + mean
```


## Mean, median, variance, sd

**Template:**

```r
mean(DATA$VAR)
median(DATA$VAR)
var(DATA$VAR)
sd(DATA$VAR)
```

**With Missing Data:**

```r
mean(DATA$VAR, na.rm = TRUE)
median(DATA$VAR, na.rm = TRUE)
var(DATA$VAR, na.rm = TRUE)
sd(DATA$VAR, na.rm = TRUE)
```

**Example:**

```{r}
mean(trashwheel$PlasticBottles, na.rm = TRUE)
median(trashwheel$PlasticBottles, na.rm = TRUE)
var(trashwheel$PlasticBottles, na.rm = TRUE)
sd(trashwheel$PlasticBottles, na.rm = TRUE)
```




## Quantiles

**Template:**

```{r}
quantile(DATA$VAR, probs = c(0.25, 0.5, 0.75))
```

**Example:**

```{r}
quantile(trashwheel$PlasticBottles, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
```


# Data visualization

## Histograms

A **histogram** shows the distribution by binning values and counting how many fall in each bin.

**Template (choose either bins OR binwidth):**

```r
ggplot(DATA, aes(x = VAR)) +
  geom_histogram(bins = X) 
```

**Example:**

```{r}
ggplot(trashwheel, aes(x = PlasticBottles)) +
  geom_histogram(bins = 20)
```

## Kernel density plot

**Template:**

```r
ggplot(DATA, aes(x = VAR)) +
  geom_density()
```

**Example:**

```{r}
ggplot(trashwheel, aes(x = PlasticBottles)) +
  geom_density()
```

## Box plots

A **box plot** summarizes median, quartiles, and potential outliers.

**Template:**

```r
ggplot(DATA, aes(VAR)) +
  geom_boxplot()
```

**Example:**

```{r}
ggplot(trashwheel, aes(PlasticBottles)) +
  geom_boxplot() 
```

## Dot plots

A **dot plot** stacks dots within bins to show distribution.

**Template:**

```r
ggplot(DATA, aes(x = VAR)) +
  geom_dotplot(binwidth = X)  # choose a sensible binwidth
```

**Example:**

```{r}
ggplot(trashwheel, aes(x = PlasticBottles)) +
  geom_dotplot(binwidth = 100)
```

# Scatter plots (two numerical variables)

A **scatter plot** reveals association, trend direction, and form.

**Template:**

```r
ggplot(DATA, aes(x = VAR1, y = VAR2)) +
  geom_point()
```

**Example:**

```{r}
ggplot(trashwheel, aes(x = PlasticBottles, y = PlasticBags)) +
  geom_point()
```

**Add a trend line (optional):**

```{r}
ggplot(trashwheel, aes(x = PlasticBottles, y = PlasticBags)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE)
```

# Quick troubleshooting

- **Lots of NAs?** Add `na.rm = TRUE` where available or filter rows with `drop_na(VAR)`.

- **Histogram looks too blocky/smooth?** Tune `bins` or `binwidth`.

- **Weird axis or units?** Check for unit conversions or outliers dominating the scale.

- **Outliers change the mean a lot.** Consider reporting **median** or using **robust** summaries.


# Appendix: minimal templates (copy‑paste)

Each template below has **placeholders** in ALL CAPS (e.g., `DATA`, `VAR`, `VAR1`, `VAR2`). Replace them with your dataset name and variable names.

```r
# Mean / Median / Var / SD
mean(DATA$VAR)
median(DATA$VAR)
var(DATA$VAR)
sd(DATA$VAR)
```
```r
# Quantiles
quantile(DATA$VAR, probs = c(0.25, 0.5, 0.75))
```
```r
# One-liner (csucistats)
num_stats(DATA$VAR)
```
```r
# Histogram
ggplot(DATA, aes(VAR)) + 
    geom_histogram(bins = X)
```
```r
# Density
ggplot(DATA, aes(VAR)) + 
    geom_density()
```
```r
# Box plot
ggplot(DATA, aes(VAR)) + 
    geom_boxplot() 
```
```r
# Dot plot
ggplot(DATA, aes(x = VAR)) + 
    geom_dotplot(binwidth = X)
```
```r
# Scatter
ggplot(DATA, aes(x = VAR1, y = VAR2)) + 
    geom_point()
```
```r
# Scatter + trend line
ggplot(DATA, aes(x = VAR1, y = VAR2)) + 
    geom_point() + 
    geom_smooth(method = "lm", se = TRUE)
```
